An Efficient Forecasting Approach\\ to Reduce Boundary Effects in Real-Time Time-Frequency Analysis

### 2021-01-27 Paper/response/response.tex

\item
\textit{page 7, second column. The authors wrote that the choice of the $M$ parameter is especially crucial when the the deterministic component of the signal is no longer stationary, like in the AHM model.
However, when they present an application of the proposed method to a signal satisfying the AHM model, neither they discuss how they select the $M$ value, nor they provide the analysis of the dependency of the performance of the proposed algorithm on the M value. Please include in section IV-C2 the evolution of the performance of the proposed method as a function of M and an explanation on how to tune its value in applications.}\\
\textbf{We have added the new paragraph IV-C2a where we discuss the influence of the parameter $M$ on the forecasting performance of an adaptive harmonic signal. In particular we mention that choosing a too small value for $M$ does not provide enough information to forecast the signal satisfactorily, while a value of $M$ that is too large makes the algorithm insensitive to local nonstationarities produced by frequency variations. The user must then make a compromise on the choice of $M$ in order to optimize the performance of the algorithm. A choice of $M$ such that the subsignals contain about a

@@

cardiac component, knowm as the cardiogenic artifact. We then show the ability of our algorithm to work on
signals containing several nonstationary components. Another application to an atrial blood pressure signal is provided in section III of the Supplementary Materials. This signal also contains both simultaneously a cardiac component and a respiratory component.}
\end{enumerate}

\section*{Reviewer 2}
Thanks for your careful evaluation of the paper and your comments, which we tried to take into account as much as we could. You will find below answers to your remarks, suggestions and questions.
\begin{enumerate}[1)]
\item
\textit{The approach is similar to the shadowing approach suggested by C. Gregobi in the Nonlinear Dynamics community, (see his Physics Review Letters in 1990 Shadowing of physical trajectories in chaotic dynamics) and subsequent papers on shadowing in multiple domains e.g. by Vargas et al in IEEE Transactions on Audio, Speech and Language in 2011.}\\
\textbf{We have mentioned this approach in the introduction of our article. We precise that dynamic mode predictors have been proposed in speech processing by Vargas and McLaughlin, to forecast signals  falling into the so-called source--filter model. This gradient-based technique rely on the shadowing approach proposed in Grebogi et al. Nevertheless,  since the source--filter is not adapted to model physiological signals on which we focus, we do not implement this method in our article.}
\item
\textit{The signal is assumed to be deterministic with additive noise, in my experience oscillatory signals of interest in many applications are a combination of deterministic and stochastic components. How robust is the scheme?}\\
\textbf{A new paragraph has been added to precise the limitations of the theoretical justifications on the performance of \textsf{SigExt}. Indeed, the models of sections III-A and III-C consider the meaningful part of the signal to be a deterministic component. These models are purposely adapted to signals showing local line spectra. The physiological signals we are interested in, such as respiratory or cardiac signals, have this characteristic (see section IV-D). Signals with wider local spectra, such as electroencephalography signals, do not fall into this category, and are more faithfully modeled as random signals. Thus, the theoretical justifications proposed above are no longer applicable to guarantee the forecasting quality of \textsf{SigExt}. A more robust approach to the study of signals following this model is envisaged in our future work.}
\item
\textit{The reference to Takens is slightly confusing could this be explained more clearly. You are not referring to nonlinear dynamic signals so I am not sure it is relevant.} \\
\textbf{Reviewer's comment is judicious. The paragraph comparing the linear dynamical model that we propose with the Takens' embedding wasn't appropriate. For the sake of clarity, this paragraph has been removed.}
\item
\textit{The paper discusses asymptotic behaviour of the forecasting algorithm but the time series lengths in practice are quite short so what impact does that have on performance.} \\
\textbf{In practice, the asymptotic behavior of \textsf{SigExt}, illustrated on a synthetic signal in the paper, can be generalized to the study of real-life signals, as long as the duration of the recorded signal largely exceeds that of the desired extension, which allows the user to set $K \gg L$. In other words, to ensure the relevance of the extension produced by \textsf{SigExt}, the user should be able to use the previously recorded signal for at least three times the duration of the desired extension, approximately. For example, in section IV-D2, the analysis of the PPG signal requires 7-second-long extensions to effectively reduce boundary effects in the TF domain. About twenty seconds of recording are thus necessary to ensure the fidelity of our algorithm.}
\end{enumerate}

### 2021-01-27 Paper/BoundEffRed.tex
paper
fig

Windowing is a common method for performing local analysis. Among many others, STFT~\cite{Flandrin:1999}, continuous wavelet transform (CWT)~\cite{Da1992}, synchrosqueezing transform (SST) \cite{Daubechies11synchrosqueezed}, and reassignment~\cite{Auger13time} (RS) are representations that fall back on the use of an analysis window. Let $x:I\to\RR$ denote the observed signal, where $I$ denotes the finite interval where the signal is measured. Let $g_s:\RR\to\RR$ denote the analysis window, where $s$ is a shape parameter. The support of $g_s$ is localized around the origin and is small with respect to $|I|$. The translation operator is $T_\tau$ defined as:
\[
T_\tau f = f(t-\tau)\ ,\quad \forall\ f:\RR\to\RR\ .
\]
Then, the local analysis of $x$ around the instant $\tau\in I$ rely on the evaluation of the following inner product:
\begin{equation}
V_x(s,\tau) = \langle x, T_\tau g_s \rangle_I \ .
\label{eq:windowing}
\end{equation}
A major shortcoming of this technique occurs when analyzing the signal $x$ {\em near the boundaries} of the interval $I$. Clearly, at these points, half of the information is missing. Consequently, the results of the inner product~\eqref{eq:windowing} are distorted. This phenomenon is usually understood as the \emph{boundary effect}. The result of the SST of a PPG is shown in the lower-left corner of Fig.~\ref{fig:ex.intro} (see section~\ref{ssse:ppg} for a comprehensive description). The distortion resulting from boundary effects is clearly visible on the right side of this representation. Indeed, while in the major left part of the image, clear lines stand out, they become blurred as they approach the right boundary of the image. Therefore, estimations of signal characteristics, like instantaneous frequencies~\cite{Delprat92asymptotic} or amplitudes, from this TF representation appear to be imprecisely determinable (or even likely to fail) in the vicinity of the boundaries. Moreover, this boundary effect would unavoidably limit applying these window-based TF analysis tools for real-time analysis purposes. It is thus desirable to have a solution to eliminating the boundary effects.

\begin{figure}
\centering
\includegraphics[width=.48\textwidth]{PPGsig.eps}
%\includegraphics[width=.24\textwidth]{zoomSSTintro.eps}
%\includegraphics[width=.24\textwidth]{zoomSSTBoundEffRed.eps}
\caption{A segment of PPG signal (top) and the right boundary of a TF representation determined by the SST without extension (bottom left), and the right boundary of a TF representation determined by the SST with the proposed boundary effect reduction algorithm by forecasting (bottom right). The window length for the SST is 12~seconds.}
\label{fig:ex.intro}
\end{figure}

>>>>

Windowing is a common method for performing local analysis. Among many others, STFT~\cite{Flandrin:1999}, continuous wavelet transform (CWT)~\cite{Da1992}, synchrosqueezing transform (SST) \cite{Daubechies11synchrosqueezed}, and reassignment~\cite{Auger13time} (RS) are representations that fall back on the use of an analysis window. Let $x:I\to\RR$ denote the observed signal, where $I$ denotes the finite interval where the signal is measured. Let $g_s:\RR\to\RR$ denote the analysis window, where $s$ is a shape parameter. The support of $g_s$ is localized around the origin and is small with respect to $|I|$. The translation operator is $T_\tau$ defined as:
\[
T_\tau f = f(t-\tau)\ ,\quad \forall\ f:\RR\to\RR\ .
\]
Then, the local analysis of $x$ around the instant $\tau\in I$ rely on the evaluation of the following inner product:
\begin{equation}
V_x(s,\tau) = \langle x, T_\tau g_s \rangle_I \ .
\label{eq:windowing}
\end{equation}
A major shortcoming of this technique occurs when analyzing the signal $x$ {\em near the boundaries} of the interval $I$. Clearly, at these points, half of the information is missing. Consequently, the results of the inner product~\eqref{eq:windowing} are distorted. This phenomenon is usually understood as the \emph{boundary effect}. The result of the SST of a PPG is shown in the lower-left corner of Fig.~\ref{fig:ex.intro} (see section~\ref{ssse:ppg} for a comprehensive description). The distortion resulting from boundary effects is clearly visible on the right side of this representation---this area is highlighted by the red dashed rectangle. Indeed, while in the major left part of the image, clear lines stand out, they become blurred as they approach the right boundary of the image. Therefore, estimations of signal characteristics, like instantaneous frequencies~\cite{Delprat92asymptotic} or amplitudes, from this TF representation appear to be imprecisely determinable (or even likely to fail) in the vicinity of the boundaries. Moreover, this boundary effect would unavoidably limit applying these window-based TF analysis tools for real-time analysis purposes. It is thus desirable to have a solution to eliminating the boundary effects.

\begin{figure}
\centering
\includegraphics[width=.48\textwidth]{PPGsig.eps}
\caption{A segment of PPG signal (top) and the right boundary of a TF representation determined by the SST without extension (bottom left), and the right boundary of a TF representation determined by the SST with the proposed boundary effect reduction algorithm by forecasting (bottom right). The window length for the SST is 12~seconds. A video of these real-time SSTs is available online at \url{https://github.com/AdMeynard/BoundaryEffectsReduction}}
\label{fig:ex.intro}
\end{figure}

### 2021-01-27 Paper/results.tex
paper

\paragraph{Influence of the subsignals length $M$}
To evaluate the forecasting quality, {\sf SigExt} is applied to $1000$ realizations of the above-described signal. We forecast signal extensions of $0.1$~s, \ie~$L=700$. Table~\ref{tab:mse.sine} shows the resulting mean and standard deviation (SD) of the experimental MSE, according to four different values of the subsignals length $M$. Note that choosing a too small $M$ does not provide enough information to forecast the signal satisfactorily. Furthermore, a value of $M$ that is too large makes the algorithm insensitive to local nonstationarities produced by frequency variations. The user must then make a compromise on the choice of $M$ in order to optimize the performance of the algorithm. A choice of $M$ such that the subsignals contain about a dozen oscillations is generally appropriate.

@@

\begin{table}
\centering
\caption{AHM Signal: Performance of the Extension Methods and the Associated Boundary-Free STFT}
\begin{tabular}{|c||c|c||c|}
  \hline
   \multirow{2}{60pt}{\centering Extension method}  & \multirow{2}{30pt}{\centering CPU time (s)} & \multirow{2}{50pt}{\centering MSE\\ (mean $\pm$ SD)} & \multirow{2}{50pt}{\centering Index $D$ of the STFT}\\
    &  & & \\
   \hhline{|=#=|=#=|}
   {\sf SigExt ($M=100$)} & $0.008$ & $1.133 \pm  0.077$ & $0.0091 \pm  0.0019$\\
   \hline
   {\sf SigExt ($M=750$)} & $0.122$ & $0.479 \pm  0.166$ & $0.0056 \pm  0.0019$\\
   \hline
   {\sf SigExt ($M\!=\!1500$)} & $0.634$ & $0.907 \pm  0.377$ & $0.0065 \pm  0.0023$\\
   \hline
   {\sf SigExt ($M\!=\!3000$)} & $2.881$ & $5.365 \pm  3.919$ & $0.0117 \pm  0.0038$\\
   \hline
   Symmetric &  $<0.001$ & $5.452 \pm  0.002$ & $1.8209 \pm  0.0025$\\
   \hline
   EDMD & $1.587$ & $0.983  \pm  0.004$ & $0.1145 \pm  0.0018$\\
   \hline
   GPR & $515.787$ & $0.479  \pm  0.166$ & $0.0119 \pm  0.0037$\\
   \hline
   TBATS  & $1776.778$ & $0.686 \pm  0.467$ & $0.3847 \pm  0.1397$\\
   \hline
\end{tabular}
\label{tab:mse.sine}
\end{table}

@@

Then, we consider a second respiratory signal that contains not only the respiratory cycle, but also the cardiogenic artifact which reflects the hemodynamic component. The top of Fig.~\ref{fig:resp.2} displays the ordinary SST and the boundary-free SST of a 180-second-long extract of this signal, sample at $125$~Hz. This TF analysis brings out both components---the respiratory component, the most energetic, is located around the fundamental frequency $0.25$~Hz and its multiples, while the cardiac component, less visible, is located around $1.7$~Hz. Note that, here, boundary effects are only reduced on the right-side boundary of the TF domain. The middle of Fig.~\ref{fig:resp.2} show a zoom near the right-boundary of the ordinary SST, while the bottom of Fig.~\ref{fig:resp.2} respectively show the respective zoom on the boundary-free SST. In this area, the boundary-free SST makes it possible to disentangle the components contained in the signal. Moreover, the performance index of this representation takes the value $D=0.314$. This means that {\sf BoundEffRed} has reduced the right-side boundary effects by about $69\%$ with respect to the ordinary SST. This shows the ability of our algorithm to work on signals containing several nonstationary components.

\begin{figure}
\centering
\includegraphics[width=.48\textwidth]{Resp2comp.eps}
\includegraphics[width=.48\textwidth]{Resp2compZOOM.eps}
\caption{Ordinary SST (top-left) and boundary-free SST (top-right) of a two-component respiratory signal. Middle and bottom: zoom on the areas delimited by the red dashed rectangle; these images are rotated 90 degrees clockwise.}
\label{fig:resp.2}
\end{figure}

>>>>

\paragraph{Influence of the subsignals length $M$}
To evaluate the forecasting quality, {\sf SigExt} is applied to $1000$ realizations of the above-described signal. We forecast signal extensions of $0.1$~s, \ie~$L=700$. Table~\ref{tab:mse.sine} shows the resulting mean and standard deviation (SD) of the experimental MSE, according to three different values of the subsignals length $M$. Note that choosing a too small $M$ does not provide enough information to forecast the signal satisfactorily. Furthermore, a too large $M$ makes the algorithm insensitive to local nonstationarities produced by frequency variations. The user must then make a compromise on the choice of $M$ in order to optimize the performance of the algorithm. A choice of $M$ such that the subsignals contain about a dozen oscillations is generally appropriate.

@@

\begin{table}
\centering
\caption{AHM Signal: Performance of the Extension Methods and the Associated Boundary-Free STFT}
\begin{tabular}{|c||c|c||c|}
  \hline
   \multirow{2}{60pt}{\centering Extension method}  & \multirow{2}{30pt}{\centering CPU time (s)} & \multirow{2}{50pt}{\centering MSE\\ (mean $\pm$ SD)} & \multirow{2}{50pt}{\centering Index $D$ of the STFT}\\
    &  & & \\
   \hhline{|=#=|=#=|}
   {\sf SigExt ($M=100$)} & $0.008$ & $1.133 \pm  0.077$ & $0.0091 \pm  0.0019$\\
   \hline
   {\sf SigExt ($M=750$)} & $0.122$ & $0.479 \pm  0.166$ & $0.0056 \pm  0.0019$\\
   \hline
   {\sf SigExt ($M\!=\!1500$)} & $0.634$ & $0.907 \pm  0.377$ & $0.0065 \pm  0.0023$\\
   \hline
%   {\sf SigExt ($M\!=\!3000$)} & $2.881$ & $5.365 \pm  3.919$ & $0.0117 \pm  0.0038$\\
%   \hline
   Symmetric &  $<0.001$ & $5.452 \pm  0.002$ & $1.8209 \pm  0.0025$\\
   \hline
   EDMD & $1.587$ & $0.983  \pm  0.004$ & $0.1145 \pm  0.0018$\\
   \hline
   GPR & $515.787$ & $0.479  \pm  0.166$ & $0.0119 \pm  0.0037$\\
   \hline
   TBATS  & $1776.778$ & $0.686 \pm  0.467$ & $0.3847 \pm  0.1397$\\
   \hline
\end{tabular}
\label{tab:mse.sine}
\end{table}

@@

Then, we consider a second respiratory signal that contains not only the respiratory cycle, but also the cardiogenic artifact which reflects the hemodynamic component~\cite{Smith94recognition}. The top of Fig.~\ref{fig:resp.2} displays the ordinary SST and the boundary-free SST of a 180-second-long extract of this signal, sample at $125$~Hz. This TF analysis brings out both components---the respiratory component, the most energetic, is located around the fundamental frequency $0.25$~Hz and its multiples, while the cardiac component, less visible, is located around $1.7$~Hz. Note that, here, boundary effects are only reduced on the right-side boundary of the TF domain. The middle of Fig.~\ref{fig:resp.2} show a zoom near the right-boundary of the ordinary SST, while the bottom of Fig.~\ref{fig:resp.2} respectively show the respective zoom on the boundary-free SST. In this area, the boundary-free SST makes it possible to disentangle the components contained in the signal. Moreover, the performance index of this representation takes the value $D=0.314$. This means that {\sf BoundEffRed} has reduced the right-side boundary effects by about $69\%$ with respect to the ordinary SST. This shows the ability of our algorithm to work on signals containing several nonstationary components.

\begin{figure}
\centering
\includegraphics[width=.48\textwidth]{Resp2comp.eps}
\includegraphics[width=.48\textwidth]{Resp2compZOOM.eps}
\caption{Ordinary SST (top-left) and boundary-free SST (top-right) of a two-component respiratory signal. Middle and bottom: zoom on the areas delimited by the red dashed rectangles; these images are rotated 90 degrees clockwise.}
\label{fig:resp.2}
\end{figure}
